---
title: "MMG3002Y pre-class assessment"
output: html_document
date: '2024-08-06'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## MMG3002Y self-assessment

This self assessment is intended to help you determine if you should do some summer prep work for the Biostatistics course. Please complete the questions below. You don’t have to return this quiz to us; however, if there are questions on this self-assessment which you don’t know how to answer or concepts you are unfamiliar with, we strongly recommend you do some prep work before the semester starts. Please contact Dr. Kaitlin Laverty - kaitlin.laverty@mail.utoronto.ca if you have any questions about the content of this self-assessment or general questions about the course. 

##1 Variables - Write a line of code to assign the value 3 to the variable "my_lucky_number"
```{r}
my_lucky_number <- 3

```

##2 Vectors - A vector is a sequence of data of the same type. Write a line of code to create a vector containing the numbers 10, 20, 30, 40, 50. Name this vector "my_vector". Note, a vector can be created in R using the c() function
```{r}
my_vector <- c(10, 20, 30, 40, 50)

```


##3 Indexing - Indexing is used for accessing subsets of vectors. Write a line of code to access the 3rd element in my_vector.
```{r}
my_vector <- [3]

```

##5a Conditional statements - Explain in words what the code below is doing:

```{r}
x = 5
if (x > 0) {
  print("Positive number")
}
```


##5b Conditional statements - Write a conditional statement which will print "negative number" if x is less than 0
```{r}

```


##6 For loops - Explain in words what the code below is doing:

```{r}
for (x in c(1,3,5)){
  print(log(x))
}
```




