---
title: "R Data Analytics"
output: html_document
date: "2024-10-24"
---

Materials and Practice
Project: Install RStudio
6 of 11 · Installing Tidyverse Packages Collection

Tidyverse is a collection of packages needed for data analysis.
These are the most common packages:
```{r}
readr #for data import.
ggplot2 #for data visualization.
dplyr #for data manipulation.
tidyr #for data cleaning.
purrr #for functional programming.
tibble #for tibbles, a modern re-imagining of dataframes.
stringr #for strings.

# install tidyverse
install.packages("tidyverse")

library(readr)
library(ggplot2)

# it's also possible to load them all using:
library(tidyverse)

```

7 of 11 · Importing Data
Learn
When we write code to import a dataset into RStudio, we must make sure that we have the working directory set to the location where our data files are stored. Otherwise, we'll have to specify the file location.

There are a few ways to set the working directory in RStudio.

Use the setwd() function. We can use this function to set our working directory to the Desktop folder by typing: setwd("~/Desktop") in the Console.

Navigate to and choose the working directory from the Session menu.

To print our working directory in the Console, type getwd().

In other lessons, we learned we'll often want to import data into R. We learned to use the read_csv() function in the readr package to import data and assign the dataframe to a variable. Datasets we import into R as dataframes display in the Environment tab as well. To demonstrate this, you can download a dataset. We will use it in the following screens.

Make sure to download this dataset and store it in your working directory.

As an alternative to typing code in the Console, we can import datasets into RStudio using the Import Dataset feature in the Environment tab.

We recommend using the readr option, which uses functions, like read_csv(), that we learned about earlier in this course.

If we import data into R using the Import Dataset feature, the data automatically displays in a tab located on the top-left of the interface.

If we import our data into R by typing functions like read_csv() in the Console, R does not automatically open the data file. To display our data, we use the function View(). For example, to look at the dataframe we created in a previous lesson when we imported data on Monster Jobs into R, we would type the following code into the Console:
```{r}
View(monster_jobs)
```

8 of 11 · Writing Scripts
In RStudio, we can write scripts in the text editor window at the top left of the interface. To create a new script, use the create new script drop down menu at the top right of the interface or commands in the file menu. You can also use the keyboard shortcut Ctrl + Shift + N for windows and Command + Shift + N for macOS.
```{r}
# loading packages
library(readr)
library(ggplot2)

# loading the monster_jobs dataset
monster_jobs_clean <- read_csv("monster_jobs_clean.csv")
view(monster_jobs_clean)

# plotting job_id vs salary_min
ggplot(data = monster_jobs_clean, aes(x = job_id, y = salary_min, color = job_type)) + geom_point() + labs(tittle = "Job ID vs Minimum Salary", x = "Job ID", y = "Minimum Salary", color = "Job Type")
```
To run a line of code we typed into our console's script, we place the cursor on the line and then, we can either click run at the top right of the script or use the following keyboard commands:

OS X: Cmd + Enter
Windows and Linux: Ctrl + Enter

We can also highlight multi-line chunks of our code that we want to run.

Sometimes, we'll write a script we want to run all at once. To highlight all the code in our script, we can use the command:

OS X: Cmd + a
Windows and Linux: Ctrl + a

When writing a script, it's good practice to begin at the top by writing code to load the packages we'll need to run the script:

```{r}
library(readr)
library(ggplot2)
```


Vectors in R
1 of 11 · Introduction to Data Structures: Generating Vectors

Learn
Understanding data structures is fundamental in R programming. They are the tools that help us store and work with data efficiently.

In this course we will learn the following data structures:

Vector: A one-dimensional container for storing values of SAME TYPE.
Matrix: A two-dimensional structure for storing values of SAME TYPE.
Lists: A versatile multi-dimensional stucture for storing values of ANY DATA TYPE/OBJECT.
Dataframe: Another two-dimensional structure for storing values of ANY DATA TYPE/OBJECT.
Here's a visual representation to help you picture these structures:


Learning about data structures involves:
- Creating them.
- Accessing their elements (indexing).
- Adding or removing elements.
- Using them in various operations and tasks.
In this lesson, we will focus on vectors. We'll build on the skills we learned in the previous course as we dive deeper into creating, manipulating, and performing calculations with vectors.

So, what is a vector? It is a data structure that holds a collection of values of the same type, allowing us to perform operations on them collectively. Think of it like an array in other languages like Python.

In the previous course, we encountered basic data types like logical, numeric (integer and double), and character. In this lesson, we will create vectors for each of these types:
- Logical vector.
- Integer vector.
- Double vector.
- Character vector.

Creating vectors is straightforward. We can generate a sequence of values in different ways:

A range of values like 1, 2, 3, and 4, can be generated as a vector using a colon (:) in R.

```{r}
a_range_vector <- 1:4
```
A sequence of values based on specific rules, such as 2, 5, and 8, can be generated using the seq() function:

```{r}
seq_2_10 <- seq(from = 2, to = 10, by = 3)
print(seq_2_10)
```
We can also repeat a value a specified number of times (times) with the rep() function:

```{r}
rep_4 <- rep(5, times = 4)
print(rep_4)
```

Plus, the rep() function can even repeat an entire vector:
```{r}
rep_6 <- rep(1:3, times = 2)
print(rep_6)
```


```{r}
vector_1 <- rep(4, times = 10)
vector_2 <- 1:12
vector_3 <- seq(from = 2.0, to = 5.50, by = 0.25)
```
2 of 11 · Creating a Vector Using Known Values
Learn
On this screen, we learn how to create a vector using known values.

A vector is created using the function c()(which stands for concatenate). This function receives a sequence of values of the SAME TYPE separated by commas (,) and it will transform them into a vector. Here are some practical examples:

A logical vector for making validation decisions in classes:
```{r}
c(TRUE, FALSE, FALSE)
c(TRUE, FALSE, TRUE, FALSE)
```

An integer vector that contains the scores of a game:
```{r}
c(12L, 3L, 24L, 8L)
c(21L, 13L)
```

A double vector that contains the selling prices of products:
```{r}
c(12.10, 24.90, 4.99)
c(21.10, 25.50, 35.50)
```

A character vector that contains names of the days of the week:
```{r}
c("Monday", "Tuesday", "Wednesday","Thursday", "Friday", "Saturday","Sunday")
c("Haruto", "Junkyu", "Asahi", "Doyoung", "Yoshi", "Jihoon")
```

Remember that R follows specific syntax rules in order for the computer to perform our instructions. If we try to store a sequence of values without the c() function:
```{r}
# selling_prices <- 12.10, 24.90, 4.99
```
We will recieve the following error message

Practice:
```{r}
math_grades <- c(92L, 87L, 85L)
exam_grades <- c(92, 90, 84, 95, 77, 92, 85)
is_stem_classes <- c(TRUE, TRUE, FALSE, FALSE, FALSE, FALSE, FALSE)
```


3 of 11 · Indexing Vectors by Position
Learn
On this screen, we'll investigate how a vector is structured and how we can index (i.e. access) an element in a vector.

Recall the vectors we created on the previous exercise:
```{r}
math_grades <- c(92L, 87L, 85L)
exam_grades <- c(92, 90, 84, 95, 77, 92, 85)
is_stem_classes <- c(TRUE, TRUE, FALSE, FALSE, FALSE, FALSE, FALSE)
```

When you store these vectors in R, each value within them is assigned a position number (called an index) in memory. These positions allow us to access individual values within the vector. R follows a 1-indexed system, meaning the first element in a vector has a position of one.

To extract values from the vector, we specify their position within square brackets ([]). For example, to retrieve the value in the third position of math_grades, we can write:
```{r}
math_grades[3]
```
which returns:
```{r}
# output: [1] 85
```
We can also extract multiple values of a vector by specifying a range of positions using a colon (:). To select the first through fourth elements of the exam_grades vector, we would write:
```{r}
exam_grades[1:4]
# output: [1] 92 90 84 95
```
What if we want to select vector elements that are not next to each other? To do so, we can create a new vector of the desired indices and use this new vector of indices to extract the desired values from the original vector. To select elements in the first, third, and seventh positions of the exam_grades vector, we would write:
```{r}
exam_grades[c(1, 3, 7)]
# output: [1] 92 84 85
```

Practice:
```{r}
math_grades <- c(92L, 87L, 85L)
exam_grades <- c(92, 90, 84, 95, 77, 92, 85)
# Add your code below
math_exam_grade <- math_grades[1]
desired_exam_grades <- exam_grades[c(2, 3, 6)]
non_stem_exam_grades <- exam_grades[3:7]
```

4 of 11 · Indexing Vectors using Logicals
Learn
On the previous screen, we learned how to index element in a vector based on the vector structure.

Here are the vectors we created in the previous exercise.
```{r}
math_exam_grade <- math_grades[1] # extracting a single element
non_stem_exam_grades <- exam_grades[c(3:7)] # extracting a range of elements
desired_exam_grades <- exam_grades[c(2, 3, 6)]  # extracting a list of desired elements
```

Let's extract the grades of STEM classes from the exam_grades vector. To guide us, the is_stem_classes vector contains the following values:
```{r}
is_stem_classes <- c(TRUE, TRUE, FALSE, FALSE, FALSE, FALSE, FALSE)
```

Based on what we have learned so far, we can create a vector of indices where is_stem_classes is TRUE and use it to extract the desired values.
```{r}
exam_grades[c(1, 2)]
# output: [1] 92 90
```

However, this approach is not scalable, especially with larger vectors where identifying every TRUE value would be cumbersome. Luckily, R offers a more elegant solution: logical indexing. Thus, to extract the grades of STEM classes using the is_stem_classes vector, we can do:
```{r}
exam_grades[is_stem_classes]
# output: [1] 92 90
```

Practice:
```{r}
is_stem_classes <- c(TRUE, TRUE, FALSE, FALSE, FALSE, FALSE, FALSE)
# Add your code below
homework_grades <- c(87, 81, 95, 86, 85, 90, 88)
preferred_homework_grades <- homework_grades[is_stem_classes]
```

Vectors in R
5 of 11 · Modifying Elements in a Vector
Learn
On the two previous screens, we learned how to extract elements from a vector using various methods.

- To extract a single element: math_grades[1].
- For a range of elements: exam_grades[c(3:7)].
- To select specific elements using an integer vector of indices: exam_grades[c(2, 3, 6)].
- For filtering elements based on a logical vector: exam_grades[is_stem_classes].

On this screen, we'll learn how to modify the elements of a vector. Luckily, we can achieve this by using two notions that we have already learned: assignment with <- and indexing.
```{r}
math_grades <- c(92L, 87L, 85L)
exam_grades <- c(92, 90, 84, 95, 77, 92, 85)
```

To replace the third grade in math_grades by 90L, we can write:
```{r}
math_grades[3] <- 90L #modify the third element of the vector
print(math_grades) 
# output: [1] 92 87 90
```

We can see that the third value has changed. Here's how R handles it:

First, it identifies the element we would like to modify using its index.
Then, it updates that specific element with the new value.
We can therefore make similar modifications using the other indexing methods.

Using a vector of indices:
```{r}
exam_grades[c(3:7)] <- 85
print(exam_grades)
# output: [1] 92 90 85 85 85 85 85
```

Using logicals
```{r}
exam_grades[is_stem_class] <- 95 
print(exam_grades) 
# output: [1] 95 95 84 95 77 92 85
```

Practice:
```{r}
math_grades <- c(92L, 87L, 85L)
exam_grades <- c(92, 90, 84, 95, 77, 92, 85)
is_stem_classes <- c(TRUE, TRUE, FALSE, FALSE, FALSE, FALSE, FALSE)

math_grades[2] <- 80L
exam_grades[is_stem_classes] <- 92
```

6 of 11 · Appending Elements to a Vector
Learn
It turns out that there are three additional class grades that we didn't initially consider in our dataset.

The exam_grades vector contains the following values.
```{r}
exam_grades <- c(92, 90, 84, 95, 77, 92, 85)
```

The three additional exams' grades are stored in the vector extra_exam_grades.
```{r}
extra_exam_grades <- c(75, 85, 95)
```

To add these three values to the existing vector we can manually add these three elements.
```{r}
new_exam_grades <- c(92, 90, 84, 95, 77, 92, 85, 75, 85, 95)
```

However, this approach can become quite cumbersome, especially with larger vectors. Fortunately, R provides an efficient solution using the c() function:
```{r}
new_exam_grades <- c(exam_grades, extra_exam_grades)
print(new_exam_grades)
# output: [1] 92 90 84 95 77 92 85 75 85 95
```

This code creates a new vector, new_exam_grades, by appending the elements in extra_exam_grades to the vector exam_grades.

We can also reverse this process by appending the elements in exam_grades to the vector extra_exam_grades by swapping their positions like this:
```{r}
new_exam_grades <- c(extra_exam_grades,exam_grades)
print(new_exam_grades) 
# output: [1] 75 85 95 92 90 84 95 77 92 85
```

practice:
```{r}
extra_is_stem_classes <- c(TRUE, FALSE, FALSE)
new_is_stem_classes_end <- c(is_stem_classes, extra_is_stem_classes)
new_is_stem_classes_start <- c(extra_is_stem_classes, is_stem_classes)